---
title: "Applying Drift Corrections with driftR"
author: "Andrew Shaughnessy, Christopher Prener, Elizabeth Hasenmueller"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{driftR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

In-situ water-quality monitoring instruments take a continuous measurements of a various chemical and physical parameters. The longer these instruments are allowed to stay in the field, the further the readings of the various sensors drift from the true value of the parameter. The purpose of this package is to correct water-quality monitoring data sets for instrumental drift in a reliable, reproducible method.

## Equations
The equations that are implemented were originally published as part of [Dr. Elizabeth Hasenmueller's](https://hasenmuellerlab.weebly.com) [dissertation research](http://openscholarship.wustl.edu/cgi/viewcontent.cgi?article=1584&amp=&context=etd&amp=&sei-redir=1&referer=https%253A%252F%252Fscholar.google.com%252Fscholar%253Fhl%253Den%2526as_sdt%253D0%25252C26%2526q%253Dhasenmueller%252Bwash%252Bu%2526btnG%253D#search=%22hasenmueller%20wash%20u%22) (see page 32).

### Correction Factor
The correction factor equation (implemented in `dr_factor()`) is as follows:

$\textrm{Let:}$

* ${f}_{t} = \textrm{correction factor}$
* $t = \textrm{time interval}$
* $\sum{t} = \textrm{total time}$

$$ {f}_{t} = \left( \frac { t }{ \sum { t }  }  \right) $$

### One-Point Calibration
The one-point calibration equation (implemented in `dr_correctOne()`) is as follows:

$\textrm{Let:}$

* $C = \textrm{drift corrected parameter}$
* $m = \textrm{uncorrected \textit{in-situ} measurement parameter}$
* ${f}_{t} = \textrm{correction factor}$
* ${s}_{i} = \textrm{calibration standard}$
* ${s}_{f} = \textrm{calibration standard after drift}$

$$ C =  m + {f}_{t} \cdot \left( { s }_{ i }-{ s }_{ f } \right) $$

### Two-Point Calibration 
The two-point calibration equation (implemented in `dr_correctTwo()`) is as follows:

$\textrm{Let:}$

* $C = \textrm{drift corrected parameter}$
* $m = \textrm{uncorrected \textit{in-situ} measurement parameter}$
* ${f}_{t} = \textrm{correction factor}$
* ${a}_{t} = \textrm{lower concentration calibration}$
* ${b}_{t} = \textrm{higher concentration calibration}$
* ${a}_{i} = \textrm{lower calibration standard}$
* ${b}_{i} = \textrm{higher calibration standard}$
* ${a}_{f} = \textrm{lower calibration standard after drift}$
* ${b}_{f} = \textrm{higher calibration standard after drift}$

$$ { a }_{ t } = { a }_{ i } + {f}_{t} \cdot \left( { a }_{ i }-{ a }_{ f } \right) $$
$$ { b }_{ t } = { b }_{ i } - {f}_{t} \cdot \left( { b }_{ i }-{ b }_{ f } \right) $$
$$ C=\left( \frac { m-a_{ t } }{ { b }_{ t }-{ a }_{ t }  }  \right) \cdot \left( { b }_{ i }-{ a }_{ i } \right) + { a }_{ i } $$

## Using driftR
There are five main functions in the `driftR` which each serve their own purpose in data cleaning process:

1. `dr_readSonde()` - import data from the Sonde 
2. `dr_factor()` - apply correction factors
3. `dr_cleanOne()` - one point calibration 
4. `dr_cleanTwo()` - two point calibration
5. `dr_drop()` - drop observations to account for equilibration

### Importing data
First, to import a data set, the `dr_sondeClean()` function will be utilized. The argument `file` tells the function where the data is located. The argument `defineVar` is a logical statement where if `FALSE`, the data will be imported raw with no modifications and if `TRUE`, the "units" observation will be removed and the collected data will be stored as numeric variables. Using `defineVar = TRUE` will format the date as Month/Day/Year.

```{r, eval=FALSE}
df <- dr_readSonde(system.file("extdata", "rawData.csv", package = "driftR"), defineVar = TRUE)
```

### Creating correction factors
The next step in the data cleaning process is creating correction factors using `dr_factor()`. The correction factors that are generated are used to determine how much of the total drift is experienced by each observation for the total data set. The argument `.data` is the working data frame for the correction. `corrFactor` is the name of the vairiable that will contain the correction factors.The argument `dateVar` is the date variable for the data set and `timeVar` is the time variable. The argument `format` describes the way that the date data is stored in the data frame and can be either "YMD" or "DMY". If the `dr_readSonde()` function is used to import the data, then `format = "MDY"`

```{r, eval=FALSE}
df <- dr_factor(df,
                 corrFactor = corfac,
                 dateVar= Date, 
                 timeVar = Time, 
                 format = "MDY")
```

### Cleaning the data
After creating the correction factors, we can begin cleaning the data for drift. In order to correct the data, there needs to be some measurement taken that tells the user what the instrument should be reading and what the instrument is actually reading. This step can be done with either one or two standard measurements. If one standard measurement is taken, then the function `dr_cleanOne()` will be used. The argument `.data` is the working data frame for the correction.  `sourceVar` is the variable name that you want to correct and `cleanVar` is the name of the variable that will contain the corrected data. The arguments `calVal` and `calStd` are what the instrument was reading and what it should have been reading respectively. The `factorVar` argument is the result generated from the `dr_correct()` function. 

If two standard measurements are taken, then the function `dr_clean2()` is utilized. The `calValLow` and `calValHigh` arguments are what the instrument was reading for the low and high concentration measurements respectively and the `calStdLow` and `calStdHigh` arguments are what the instrument should have been reading for the low and high concentration measurements respectively. 

```{r, eval=FALSE}
df <- dr_cleanOne(df,
                  sourceVar = SpCond, 
                  cleanVar = SpCond_corr, 
                  calVal = 1.07, 
                  calStd = 1, 
                  factorVar = corfac)

df <- dr_cleanTwo(df, 
                  sourceVar = pH, 
                  cleanVar = pH_corr,
                  calValLow = 7.01, 
                  calStdLow = 7, 
                  calValHigh = 11.8, 
                  calStdHigh = 10, 
                  factorVar = corfac)
```

### Dropping Data
After the data has been cleaned, some of the data will need to be dropped in order to account for equilibration of the sensors as well as time out of the water during calibration. It is important to do this step last because dropping data before using `dr_clean1()` or `dr_clean2()` will result in the corrections being non-consistent. To do this, the `dr_drop()` function will be used. The argument `.data` is the working data frame for the correction. The arguments `head` and `tail` are the number of observations to be dropped from the beginning and the end of the data set respectively. It is recommended that at least 30 minutes are dropped from both the beginning and end of the data set (i.e. if an observation is taken every 5 minutes, then `head = 6`and `tail = 6`).

```{r, eval=FALSE}
df <- dr_drop(df, 
              head = 6, 
              tail = 6)
```

## Tidying Data
### Reordering Variables

Variable names will need to be re-ordered after using `driftR` to get like variables next to each other in your data frame. You can use the [`select()`](http://dplyr.tidyverse.org/reference/select.html) function from [`dplyr`](http://dplyr.tidyverse.org) to accomplish this:

```{r, eval=FALSE}
df <- select(df, Date, Time, SpCond, SpCond_Corr, pH, pH_Corr, pHmV, pHmV_Corr, 
             Chloride, Chloride_Corr, AmmoniumN, AmmoniumN_Corr, NitrateN, NitrateN_Corr, 
             Turbidity, Turbidity_Corr, DO, DO_Corr)
```

Like all other `dplyr` functions, `select()` can be included in a pipe as well.

### Removing Unneeded Variables

Finally, if there are unneeded variables left in your data set at the end of the pre-processing stage, you can also use the [`select()`](http://dplyr.tidyverse.org/reference/select.html) function from [`dplyr`](http://dplyr.tidyverse.org) to remove them. The function accepts the data frame name followed by a comma and a negative sign in front of the variable to be removed. For example, the `NitrateN` variable does not contain any non-zero observations. We could remove it like so:

```{r, eval=FALSE}
df <- select(df, -NitrateN))
```

If there were multiple variables to be removed, a list of the variables can be provided inside `-c(varlist)`:

```{r, eval=FALSE}
df <- select(df, -c(variable1, variable2)))
```

## Exporting Data
Finally, data can be exported to `csv` (our recommended file format because it is plain-text and non-proprietary) using the [`readr` package](http://readr.tidyverse.org) using the [`write_csv()`](http://readr.tidyverse.org/reference/write_delim.html) function:

```{r, eval=FALSE}
write_csv(df, path = "waterData.csv", na = "NA")
```
