---
title: "Using Data from Other Sources"
author: "Andrew Shaughnessy, Christopher Prener, Elizabeth Hasenmueller"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Data from Other Sources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

While `driftR` is designed to work seamlessly with output from YSI Multiparameter V2 Sonde, YSI EXO2, and Onset U24 Conductivity Logger instruments, it can also be used to correct data from other sources. There are only a few steps that would be needed to get data into a tidy `driftR` format. Below are example data after they have been imported using the `dr_read()` function. This is the expected format that `driftR` requires, so data from other sources must be modified to this configuration. 

```{r, eval=FALSE}
# A tibble: 1,527 x 11
        Date     Time  Temp SpCond    pH  pHmV Chloride AmmoniumN NitrateN  Turbidity    DO
       <chr>    <chr> <dbl>  <dbl> <dbl> <dbl>    <dbl>     <dbl>    <int>      <dbl> <dbl>
 1 9/18/2015 12:10:49 14.76  0.754  7.18 -36.4    51.22      3.35        0        3.7 92.65
 2 9/18/2015 12:15:50 14.64  0.750  7.14 -34.1    49.62      6.29        0       -0.2 93.73
 3 9/18/2015 12:20:51 14.57  0.750  7.14 -33.9    49.75      7.84        0       -0.1 93.95
 4 9/18/2015 12:25:51 14.51  0.749  7.13 -33.9    50.32      7.67        0       -0.2 93.23
 5 9/18/2015 12:30:51 14.50  0.749  7.13 -33.6    50.74      7.13        0        0.0 92.74
 6 9/18/2015 12:35:51 14.63  0.749  7.13 -33.5    50.84      6.49        0        0.0 93.71
 7 9/18/2015 12:40:51 14.69  0.749  7.13 -33.6    50.66      5.78        0       -0.2 94.56
 8 9/18/2015 12:45:51 14.66  0.749  7.12 -33.3    50.23      5.32        0       -0.2 94.16
 9 9/18/2015 12:50:52 14.65  0.749  7.12 -33.3    50.49      4.89        0       -0.2 93.58
10 9/18/2015 12:55:51 14.69  0.749  7.12 -33.1    50.04      4.60        0       -0.2 93.80
# ... with 1,517 more rows
```

The sections below detail pre-processing steps that you may have to take to prepare your data for use with `driftR`.

## Importing Data

Data come in a variety of formats, and importing them into `R` can occasionally be a challenge.

* If your data are included in a flat file like a `csv`, `tsv`, `txt`, or another delimited file format, we recommend using the [`readr` package](http://readr.tidyverse.org).
* If your data are included in a Microsoft Excel file, we recommend using the [`readxl` package](http://readxl.tidyverse.org).
* If your data are included in a SPSS, SAS, or Stata file, we recommend using the [`haven` package](http://haven.tidyverse.org).  
* If your data are in a Microsoft Access database, we recommend using the [`RODBC` package](https://CRAN.R-project.org/package=RODBC). This will require a Windows computer, 32-bit `R`, and either Microsoft Access or the appropriate drivers installed.
* If your data are in another database format, we recommend [this article](http://db.rstudio.com/) on using Databases with `R`.

All of the example code below assumes that you have a data frame named `waterData`.

## Metadata
No metadata should be stored in the observations. If metadata are present, remove them using the following technique. (This example assumes that metadata is stored in row 1):

```{r, eval=FALSE}
waterData <- waterData[-1,]
```

If there are multiple lines of metadata, they can be removed like so:

```{r, eval=FALSE}
waterData <- waterData[-c(1, 2),]
```

## Tibbles

Given the typically large data sets for these intruments, we encourage (but do not enforce) data to be stored as [tibbles](http://tibble.tidyverse.org). Tibbles are the [`tidyverse`](https://www.tidyverse.org) implementation of data frames. They print in a more organized manner and they behave in a more stable fashion. To convert your data to a tibble, use the function [`as_tibble()`](http://tibble.tidyverse.org/reference/as_tibble.html):

```{r, eval=FALSE}
library(tibble)

waterTibble <- as_tibble(waterData)
```

## Variable Names

Variable names should be short and descriptive. We recommend using `camelCase` or `snake_case` to name variables. Use the [`rename()`](http://dplyr.tidyverse.org/reference/select.html) function from [`dplyr`](http://dplyr.tidyverse.org) to accomplish this. The function accepts the data frame name followed by a comma and the new name set equal to the old name:

```{r, eval=FALSE}
waterTibble <- rename(waterTibble, date = a.very.long.date.variable.name)
```

If you have a number of variables to rename, you can pipe them together:

```{r, eval=FALSE}
waterTibble <- waterTibble %>%
  rename(date = a.very.long.date.variable.name) %>%
  rename(time = a.very.long.time.variable.name)
```

## Specific Variables
### Date and Time
Please check out our vignette on [dates and times in driftR](DatesTimes.html) for additional details on how these should be formatted.

### Temperature

`driftR` makes no direct use of the `Temp` data included in output. The [`weathermetrics` package](https://CRAN.R-project.org/package=weathermetrics) includes functions for conversions between Celsius and Fahrenheit.

### Other Variables

Beyond date and time data, all variables should be stored as either double, integer, or numeric values:

```{r, eval=FALSE}
waterTibble$measure <- as.numeric(waterTibble$measure)
waterTibble$measure <- as.integer(waterTibble$measure)
waterTibble$measure <- as.double(waterTibble$measure)
```

### Building Pipes with Date, Time, and Other Functions

You can integrate all of the non-tidyverse functions described here into [`mutate()`](http://dplyr.tidyverse.org/reference/mutate.html) function calls. The `mutate()` function is from the [`dplyr`](http://dplyr.tidyverse.org) package.

```{r, eval=FALSE}
waterTibble <- waterTibble %>%
  mutate(temp = as.double(temp)) %>%
  mutate(pH = as.double(pH)) %>%
  mutate(NitrateN = as.integer(NitrateN))
```

## Removing Unnecessary Variables

Finally, if there are unnecessary variables left in your data set at the end of the pre-processing stage, you can use the [`select()`](http://dplyr.tidyverse.org/reference/select.html) function from [`dplyr`](http://dplyr.tidyverse.org) to remove them. The function accepts the data frame name followed by a comma and a list of the variables to be removed inside `-c(varlist)`:

```{r, eval=FALSE}
waterTibble <- select(waterTibble, -c(a.very.long.pH.variable.name, a.very.long.chloride.variable.name))
```

Like all other `dplyr` functions, `select()` can be included in a pipe as well.
